---
title: "MA 551 - Homework 3"
author: "Adam Field"
date: "March 31, 2026"
output:
  pdf_document:
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)

library(ggplot2)

set.seed(42)
```

\newpage

# Problem 1: Cauchy Distribution and Rejection Sampling

## Inversion Method for the Cauchy Distribution

The Cauchy$(\alpha)$ distribution has CDF
$$
F(x) = \frac{1}{2} + \frac{1}{\pi}\arctan\!\left(\frac{x}{\alpha}\right).
$$
Setting $U = F(X)$ and solving for $X$ gives the inversion formula
$$
X = \alpha\tan\!\bigl(\pi(U - \tfrac{1}{2})\bigr), \qquad U \sim \mathrm{Uniform}(0,1).
$$

```{r p1_cauchy_inv}
rcauchy_inv <- function(n, alpha = 1) {
  u <- runif(n)
  alpha * tan(pi * (u - 0.5))
}

# Verify against built-in Cauchy quantiles
set.seed(42)
x_test   <- rcauchy_inv(10000, alpha = 1)
ks_pval  <- ks.test(x_test, "pcauchy", location = 0, scale = 1)$p.value
cat(sprintf("KS test p-value vs Cauchy(0,1): %.4f\n", ks_pval))
```

The large $p$-value from the Kolmogorov-Smirnov test confirms that the inversion sample is
consistent with the Cauchy$(0,1)$ distribution.

## Optimal Cauchy Envelope for N(0,1) Rejection Sampling

To generate $X \sim N(0,1)$ using a Cauchy$(\alpha)$ envelope we require $\phi(x) \leq k \cdot h_\alpha(x)$
for all $x$, where $\phi(x) = (2\pi)^{-1/2}e^{-x^2/2}$ is the standard normal density and
$h_\alpha(x) = \alpha/[\pi(\alpha^2 + x^2)]$ is the Cauchy$(\alpha)$ density.  The minimum valid
scaling constant is
$$
k(\alpha) = \sup_x \frac{\phi(x)}{h_\alpha(x)}
= \sup_x \frac{\pi(\alpha^2 + x^2)\,e^{-x^2/2}}{\alpha\sqrt{2\pi}}.
$$
Differentiating the numerator with respect to $x$ and setting it to zero yields
$x\,e^{-x^2/2}(2 - \alpha^2 - x^2) = 0$, so the interior critical point satisfies $x^{*2} = 2 - \alpha^2$,
which is real for $\alpha < \sqrt{2}$.  Substituting $x^{*2} = 2 - \alpha^2$ back gives
$$
k(\alpha) = \frac{\sqrt{2\pi}\,e^{\alpha^2/2 - 1}}{\alpha}.
$$
To find the $\alpha$ that minimises $k(\alpha)$ we differentiate:
$$
\frac{\mathrm{d}}{\mathrm{d}\alpha}\left[\frac{e^{\alpha^2/2-1}}{\alpha}\right]
= \frac{e^{\alpha^2/2-1}(\alpha^2 - 1)}{\alpha^2} = 0
\implies \alpha^* = 1.
$$
At $\alpha^* = 1$ the optimal constant is
$$
k^* = \sqrt{2\pi}\,e^{-1/2} = \sqrt{\frac{2\pi}{e}} \approx 1.5203,
$$
giving a theoretical acceptance probability of $1/k^* = \sqrt{e/(2\pi)} \approx 65.8\%$.

```{r p1_optimal}
alpha_opt <- 1
k_opt     <- sqrt(2 * pi / exp(1))
cat(sprintf("Optimal alpha: %.4f\n", alpha_opt))
cat(sprintf("Optimal k:     %.6f\n", k_opt))
cat(sprintf("Theoretical acceptance probability: %.4f\n", 1 / k_opt))
```

```{r p1_rejection_sampler}
rnorm_cauchy_rej <- function(N, alpha, k) {
  samples    <- numeric(N)
  n_accepted <- 0
  n_rejected <- 0

  while (n_accepted < N) {
    x_prop <- rcauchy_inv(1, alpha)
    u      <- runif(1)
    f_x    <- dnorm(x_prop)
    h_x    <- alpha / (pi * (alpha^2 + x_prop^2))
    if (u <= f_x / (k * h_x)) {
      n_accepted <- n_accepted + 1
      samples[n_accepted] <- x_prop
    } else {
      n_rejected <- n_rejected + 1
    }
  }
  list(samples = samples, n_rejected = n_rejected)
}

set.seed(42)
res_p1 <- rnorm_cauchy_rej(1000, alpha_opt, k_opt)

cat(sprintf("Total rejections: %d\n", res_p1$n_rejected))
cat(sprintf("Empirical rejection rate: %.4f\n",
            res_p1$n_rejected / (1000 + res_p1$n_rejected)))
```

```{r p1_histogram, fig.height = 4.5}
x_grid   <- seq(-4, 4, length.out = 400)
df_theo  <- data.frame(x = x_grid, y = dnorm(x_grid))
df_samp  <- data.frame(x = res_p1$samples)

ggplot(df_samp, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 40, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_line(data = df_theo, aes(x = x, y = y),
            color = "tomato", linewidth = 1.2) +
  labs(title = "N(0,1) Sample via Cauchy Envelope Rejection Sampling (n = 1000)",
       x = "x", y = "Density") +
  theme_bw()
```

The empirical rejection rate is close to the theoretical value of approximately $34.2\%$.  The
histogram follows the superimposed $N(0,1)$ density closely, confirming that the algorithm
correctly produces standard normal samples.

\newpage

# Problem 2: Conditional Gamma Distribution

## Part (a): Density of Gamma(2,1) Conditional on X > 4

The $\mathrm{Gamma}(2,1)$ distribution has density $f_{\mathrm{Gamma}}(x) = x\,e^{-x}$ for $x > 0$.
By integration by parts,
$$
P(X > 4) = \int_4^\infty x\,e^{-x}\,\mathrm{d}x
= \Bigl[-x\,e^{-x}\Bigr]_4^\infty + \int_4^\infty e^{-x}\,\mathrm{d}x
= 4e^{-4} + e^{-4} = 5e^{-4}.
$$
The conditional density is therefore
$$
f(x) = \frac{f_{\mathrm{Gamma}}(x)}{P(X > 4)}
= \frac{x\,e^{-x}}{5\,e^{-4}}
= \frac{x}{5}\,\exp\!\bigl\{-(x-4)\bigr\}, \quad x > 4. \qquad \blacksquare
$$
One may verify that $f$ integrates to one: setting $u = x - 4$ gives
$\tfrac{1}{5}\int_0^\infty (u+4)e^{-u}\,\mathrm{d}u = \tfrac{1}{5}[\Gamma(2) + 4] = 1$.

## Part (b): Envelope Densities and Optimal Constants

### Envelope $h_1(x)$: Exponential(1/2) Conditional on X > 4

For $Y \sim \mathrm{Exp}(1/2)$ with density $\tfrac{1}{2}e^{-y/2}$,
$$
P(Y > 4) = \int_4^\infty \tfrac{1}{2}e^{-y/2}\,\mathrm{d}y = e^{-2}.
$$
The conditional density is
$$
h_1(x) = \frac{\tfrac{1}{2}e^{-x/2}}{e^{-2}} = \frac{1}{2}\,e^{-(x-4)/2}, \quad x > 4.
$$
The ratio to be maximised is
$$
\frac{f(x)}{h_1(x)} = \frac{(x/5)\,e^{-(x-4)}}{(1/2)\,e^{-(x-4)/2}} = \frac{2x}{5}\,e^{-(x-4)/2}.
$$
Its derivative equals $(2/5)\,e^{-(x-4)/2}(1 - x/2)$, which is strictly negative for $x > 4$.
Hence $f/h_1$ is strictly decreasing and its maximum is attained at $x = 4$:
$$
k_1 = \frac{2 \cdot 4}{5} = \frac{8}{5} = 1.6.
$$

### Envelope $h_2(x)$: Standard Cauchy Conditional on X > 4

For the standard Cauchy density $1/[\pi(1+x^2)]$,
$$
P(X_C > 4) = \int_4^\infty \frac{\mathrm{d}x}{\pi(1+x^2)}
= \frac{1}{\pi}\Bigl[\arctan x\Bigr]_4^\infty
= \frac{1}{\pi}\!\left(\frac{\pi}{2} - \arctan 4\right).
$$
Writing $C_2 = \pi/2 - \arctan(4)$, the conditional density is
$$
h_2(x) = \frac{1/({\pi(1+x^2)})}{C_2/\pi} = \frac{1}{C_2(1+x^2)}, \quad x > 4.
$$
The ratio to be maximised is
$$
\frac{f(x)}{h_2(x)} = \frac{(x/5)\,e^{-(x-4)}}{1/[C_2(1+x^2)]}
= \frac{C_2}{5}\,x(1+x^2)\,e^{-(x-4)}.
$$
Differentiating $x(1+x^2)e^{-(x-4)}$ with respect to $x$ yields
$e^{-(x-4)}\bigl(1 - x + 3x^2 - x^3\bigr)$.
At $x = 4$ this equals $e^0(1 - 4 + 48 - 64) = -19 < 0$, and the cubic $x^3 - 3x^2 + x - 1$
is strictly positive for $x > 4$ (so the bracketed factor is strictly negative there).
Therefore $f/h_2$ is decreasing on $(4,\infty)$ and its maximum is at $x = 4$:
$$
k_2 = \frac{C_2}{5} \cdot 4 \cdot 17 \cdot 1 = \frac{68\,C_2}{5}.
$$

```{r p2_constants}
C2 <- pi/2 - atan(4)
k1 <- 8/5
k2 <- 68 * C2 / 5

cat(sprintf("C2 = pi/2 - arctan(4) = %.6f\n", C2))
cat(sprintf("k1 = %.4f  (acceptance rate = %.4f)\n", k1, 1/k1))
cat(sprintf("k2 = %.4f  (acceptance rate = %.4f)\n", k2, 1/k2))
```

## Part (c): rej\_sample Implementation

The inverse CDF of $h_1$ follows from $H_1(x) = 1 - e^{-(x-4)/2}$:
$$
H_1^{-1}(u) = 4 - 2\log(1 - u).
$$
For $h_2$, setting $H_2(x) = [\arctan(x) - \arctan(4)]/C_2 = u$ gives
$$
H_2^{-1}(u) = \tan\!\bigl(\arctan(4) + u\,C_2\bigr).
$$

```{r p2_functions}
# Target density f(x)
fx <- function(x) ifelse(x > 4, (x / 5) * exp(-(x - 4)), 0)

# Envelope h1(x): Exp(1/2) shifted to start at 4
h1x <- function(x) ifelse(x > 4, 0.5 * exp(-0.5 * (x - 4)), 0)

# Inverse CDF of h1
ihx1 <- function(u) 4 - 2 * log(1 - u)

# Envelope h2(x): standard Cauchy conditional on X > 4
h2x <- function(x) ifelse(x > 4, 1 / (C2 * (1 + x^2)), 0)

# Inverse CDF of h2
ihx2 <- function(u) tan(atan(4) + u * C2)

# General rejection sampler (signature as specified)
rej_sample <- function(N, k, fx, hx, ihx) {
  x.sim  <- numeric(N)
  N.rej  <- 0
  n.acc  <- 0

  while (n.acc < N) {
    u1     <- runif(1)
    x.prop <- ihx(u1)
    u2     <- runif(1)
    if (u2 <= fx(x.prop) / (k * hx(x.prop))) {
      n.acc          <- n.acc + 1
      x.sim[n.acc]   <- x.prop
    } else {
      N.rej <- N.rej + 1
    }
  }
  return(list(x.sim, N.rej))
}
```

```{r p2_run}
set.seed(42)
out_h1 <- rej_sample(1000, k1, fx, h1x, ihx1)
set.seed(42)
out_h2 <- rej_sample(1000, k2, fx, h2x, ihx2)

cat("Envelope h1 (Exponential(1/2) | X > 4):\n")
cat(sprintf("  Total rejections:      %d\n",   out_h1[[2]]))
cat(sprintf("  Empirical reject rate: %.4f\n",
            out_h1[[2]] / (1000 + out_h1[[2]])))

cat("\nEnvelope h2 (Standard Cauchy | X > 4):\n")
cat(sprintf("  Total rejections:      %d\n",   out_h2[[2]]))
cat(sprintf("  Empirical reject rate: %.4f\n",
            out_h2[[2]] / (1000 + out_h2[[2]])))
```

```{r p2_plot, fig.height = 5}
x_g2 <- seq(4.01, 13, length.out = 600)

df_env <- data.frame(
  x     = rep(x_g2, 3),
  y     = c(fx(x_g2), k1 * h1x(x_g2), k2 * h2x(x_g2)),
  Curve = rep(c("f(x)", "k1 * h1(x)", "k2 * h2(x)"), each = length(x_g2))
)

ggplot(df_env, aes(x = x, y = y, color = Curve, linetype = Curve)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values    = c("black", "steelblue", "tomato")) +
  scale_linetype_manual(values = c("solid",  "dashed",    "dotdash")) +
  coord_cartesian(ylim = c(0, 0.55)) +
  labs(title    = "Target Density and Scaled Envelopes for Gamma(2,1) | X > 4",
       x        = "x",
       y        = "Density",
       color    = NULL,
       linetype = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

**Relation between rejection counts and the plot.**  A proposed draw $x$ is rejected whenever
the second uniform draw $U_2$ exceeds $f(x)/[k \cdot h(x)]$, so the overall rejection probability
equals $1 - 1/k$.  Geometrically, this corresponds to the area of the gap between the scaled
envelope $k \cdot h(x)$ and the target $f(x)$: a tighter envelope leaves less wasted area above
$f$, leading to fewer rejections.  The plot confirms that $k_1 h_1(x)$ hugs $f(x)$ far more
closely than $k_2 h_2(x)$ throughout the support, consistent with the much smaller value
$k_1 = 1.6$ compared to $k_2 \approx 3.33$ and the substantially higher rejection count produced
by the Cauchy envelope.  The Cauchy tail decays algebraically whereas $f$ decays exponentially,
so $h_2$ is a poor fit for large $x$ and wastes considerable probability mass in that region.

\newpage

# Problem 3: Monte Carlo Integration

We estimate $I = \int_0^1 \sqrt{1 - x^2}\,\mathrm{d}x$.  The exact value is the area of a unit
quarter-circle, $I = \pi/4 \approx 0.7854$.

```{r p3_true}
I_true <- pi / 4
cat(sprintf("Exact value I = pi/4 = %.8f\n", I_true))
```

## Part (a): Simple Monte Carlo

The simple Monte Carlo estimator is $\hat{I}_{\mathrm{MC}} = n^{-1}\sum_{i=1}^n g(U_i)$ where
$g(u) = \sqrt{1-u^2}$ and $U_i \stackrel{\mathrm{iid}}{\sim} U(0,1)$.

```{r p3_mc}
set.seed(42)
n    <- 100000
u_mc <- runif(n)
g_mc <- sqrt(1 - u_mc^2)

I_mc   <- mean(g_mc)
var_mc <- var(g_mc) / n
se_mc  <- sqrt(var_mc)

cat(sprintf("Simple MC estimate: %.6f\n", I_mc))
cat(sprintf("Standard error:     %.6f\n", se_mc))
cat(sprintf("Absolute error:     %.6f\n", abs(I_mc - I_true)))
```

The estimate is within one standard error of the true value, as expected.

## Part (b): Importance Sampling

We compare three importance sampling densities on $(0, 1)$.

**Density 1** $f_1(x) = 1$ (Uniform): this is equivalent to simple Monte Carlo and serves as a
baseline.

**Density 2** $f_2(x) = 2x$ (Beta$(2,1)$): concentrates mass near $x = 1$ where the integrand
$\sqrt{1-x^2}$ is small.  Inverse CDF: $F_2^{-1}(u) = \sqrt{u}$.  Importance weight:
$w_2(x) = \sqrt{1-x^2}/(2x)$.

**Density 3** $f_3(x) = (\pi/2)\cos(\pi x/2)$: largest near $x = 0$ where $\sqrt{1-x^2}$ is also
largest; the two functions have similar shapes, so the importance weight $w_3(x) = \sqrt{1-x^2}/[(\pi/2)\cos(\pi x/2)]$
is nearly constant.  Inverse CDF: $F_3^{-1}(u) = (2/\pi)\arcsin(u)$.

```{r p3_is}
set.seed(42)
n_is <- 100000

# Density 1: Uniform (identical to simple MC)
u1     <- runif(n_is)
w1     <- sqrt(1 - u1^2)
I_is1  <- mean(w1)
var_is1 <- var(w1) / n_is

# Density 2: f2 = 2x, inverse CDF x = sqrt(u)
u2      <- runif(n_is)
x2      <- sqrt(u2)
w2      <- sqrt(1 - x2^2) / (2 * x2)
I_is2   <- mean(w2)
var_is2 <- var(w2) / n_is

# Density 3: f3 = (pi/2)*cos(pi*x/2), inverse CDF x = (2/pi)*arcsin(u)
u3      <- runif(n_is)
x3      <- (2 / pi) * asin(u3)
w3      <- sqrt(1 - x3^2) / ((pi / 2) * cos(pi * x3 / 2))
I_is3   <- mean(w3)
var_is3 <- var(w3) / n_is

cat(sprintf("IS Density 1 (Uniform):   estimate = %.6f, variance = %.3e\n",
            I_is1, var_is1))
cat(sprintf("IS Density 2 (Beta(2,1)): estimate = %.6f, variance = %.3e\n",
            I_is2, var_is2))
cat(sprintf("IS Density 3 (Cosine):    estimate = %.6f, variance = %.3e\n",
            I_is3, var_is3))
cat(sprintf("\nVariance reduction relative to Uniform:\n"))
cat(sprintf("  Density 2 (Beta(2,1)): %+.2f%%\n",
            100 * (1 - var_is2 / var_is1)))
cat(sprintf("  Density 3 (Cosine):    %+.2f%%\n",
            100 * (1 - var_is3 / var_is1)))
```

Density 3 (cosine) achieves a large variance reduction because its shape closely mirrors
$\sqrt{1-x^2}$, keeping the importance weights nearly constant throughout $(0,1)$.  In contrast,
Density 2 (Beta$(2,1)$) overweights the region near $x = 1$ where $g$ is nearly zero, producing
highly variable weights and roughly no improvement over simple Monte Carlo.  These results illustrate
that the effectiveness of importance sampling depends critically on how well the proposal density
matches the shape of the integrand.

## Part (c): Stratified Sampling

We divide $(0,1)$ into $L = 3$ equal subintervals of width $h = 1/3$: $(0, 1/3)$, $(1/3, 2/3)$,
and $(2/3, 1)$.  Within each stratum $l$ we draw $m = n/L$ uniform samples and form the estimate
$\hat{I}_{\mathrm{strat}} = h \sum_l \bar{g}_l$.  The stratified variance is
$\mathrm{Var}(\hat{I}_{\mathrm{strat}}) = (h^2/m)\sum_l \sigma_l^2$, where $\sigma_l^2$ is the
within-stratum variance of $g$.

```{r p3_strat}
set.seed(42)
n_strat     <- 99999      # divisible by 3
L           <- 3
m_per       <- n_strat / L
h_width     <- 1 / L

bounds      <- seq(0, 1, length.out = L + 1)
strat_means <- numeric(L)
strat_vars  <- numeric(L)

for (l in seq_len(L)) {
  u_l           <- runif(m_per, min = bounds[l], max = bounds[l + 1])
  g_l           <- sqrt(1 - u_l^2)
  strat_means[l] <- mean(g_l)
  strat_vars[l]  <- var(g_l)
}

I_strat    <- h_width * sum(strat_means)
var_strat  <- (h_width^2 / m_per) * sum(strat_vars)
pct_strat  <- 100 * (var_mc - var_strat) / var_mc

cat(sprintf("Stratified estimate:     %.6f\n", I_strat))
cat(sprintf("Stratified variance:     %.3e\n", var_strat))
cat(sprintf("Simple MC variance:      %.3e\n", var_mc))
cat(sprintf("Estimated variance red.: %.2f%%\n", pct_strat))
```

Stratified sampling eliminates the between-stratum component of variance, which is substantial
here because $g(x) = \sqrt{1-x^2}$ decreases steeply from 1 to 0 across the unit interval.
Forcing equal-weight coverage of each subinterval ensures that the sampler explores all parts of
the domain systematically, yielding a noticeably tighter estimate than simple Monte Carlo.

## Part (d): Antithetic Sampling

For $U_i \stackrel{\mathrm{iid}}{\sim} U(0,1)$, the pair $\bigl(g(U_i),\, g(1-U_i)\bigr)$
is negatively correlated: as $U$ increases, $g(U) = \sqrt{1-U^2}$ decreases while
$g(1-U) = \sqrt{U(2-U)}$ increases.  The antithetic estimator uses $n/2$ pairs:
$$
\hat{I}_{\mathrm{anti}} = \frac{1}{n/2}\sum_{i=1}^{n/2}\frac{g(U_i) + g(1-U_i)}{2}.
$$

```{r p3_anti}
set.seed(42)
n_anti  <- 50000           # n/2 pairs gives n = 100000 function evaluations
u_anti  <- runif(n_anti)
g_pair  <- (sqrt(1 - u_anti^2) + sqrt(1 - (1 - u_anti)^2)) / 2

I_anti   <- mean(g_pair)
var_anti <- var(g_pair) / n_anti
pct_anti <- 100 * (var_mc - var_anti) / var_mc

emp_cor  <- cor(sqrt(1 - u_anti^2), sqrt(1 - (1 - u_anti)^2))

cat(sprintf("Antithetic estimate:     %.6f\n", I_anti))
cat(sprintf("Antithetic variance:     %.3e\n", var_anti))
cat(sprintf("Simple MC variance:      %.3e\n", var_mc))
cat(sprintf("Estimated variance red.: %.2f%%\n", pct_anti))
cat(sprintf("Empirical correlation:   %.4f\n", emp_cor))
```

The strong negative empirical correlation between $g(U)$ and $g(1-U)$ drives a substantial
variance reduction while halving the number of random number draws required.  Antithetic sampling
is especially effective for monotone integrands because the symmetry $U \mapsto 1 - U$ always
produces a value in the complementary region of the domain.

## Part (e): Control Variate

We approximate $g(x) = \sqrt{1-x^2}$ by the straight line through $(0,1)$ and $(1,0)$:
$c(x) = 1 - x$.  Since $E[c(U)] = E[1-U] = 1/2$ is known analytically, $c$ serves as a
control variate.  The estimator is
$$
\hat{I}_{\mathrm{CV}} = \frac{1}{n}\sum_{i=1}^n g(U_i) - \hat{b}\!\left(\frac{1}{n}\sum_{i=1}^n(1-U_i) - \frac{1}{2}\right),
$$
where the optimal coefficient $\hat{b} = \widehat{\mathrm{Cov}}(g(U), 1-U) / \widehat{\mathrm{Var}}(1-U)$
is estimated from the same sample.  The resulting variance is
$\mathrm{Var}(\hat{I}_{\mathrm{CV}}) = \mathrm{Var}(g(U) - \hat{b}(c(U) - 1/2))/n$.

```{r p3_cv}
set.seed(42)
n_cv  <- 100000
u_cv  <- runif(n_cv)
g_cv  <- sqrt(1 - u_cv^2)
c_cv  <- 1 - u_cv           # control variate

b_hat    <- cov(g_cv, c_cv) / var(c_cv)
adj_cv   <- g_cv - b_hat * (c_cv - 0.5)
I_cv     <- mean(adj_cv)
var_cv   <- var(adj_cv) / n_cv
pct_cv   <- 100 * (var_mc - var_cv) / var_mc

cat(sprintf("Estimated optimal b:     %.4f\n",  b_hat))
cat(sprintf("CV estimate:             %.6f\n",  I_cv))
cat(sprintf("CV variance:             %.3e\n",  var_cv))
cat(sprintf("Simple MC variance:      %.3e\n",  var_mc))
cat(sprintf("Estimated variance red.: %.2f%%\n", pct_cv))
```

The optimal coefficient $\hat{b}$ measures the regression slope of $g(U)$ on $c(U) = 1-U$.  The
linear approximation $1 - x$ is closely correlated with $\sqrt{1-x^2}$ over $(0,1)$, so
subtracting the correlated part of the noise yields a large variance reduction.  The analytical
value is $b^* = 12(\pi/8 - 1/3) \approx 0.71$, which the estimator recovers closely.

## Summary

```{r p3_summary}
methods    <- c("Simple MC", "IS: Uniform", "IS: Beta(2,1)", "IS: Cosine",
                "Stratified (L=3)", "Antithetic", "Control Variate")
estimates  <- c(I_mc,  I_is1, I_is2,  I_is3,  I_strat, I_anti, I_cv)
variances  <- c(var_mc, var_is1, var_is2, var_is3, var_strat, var_anti, var_cv)
pct_reds   <- 100 * (var_mc - variances) / var_mc

results_df <- data.frame(
  Method      = methods,
  Estimate    = round(estimates, 6),
  Variance    = signif(variances, 3),
  Pct_Red     = round(pct_reds, 1)
)
print(results_df, row.names = FALSE)
```

The cosine importance sampling density and the control variate approach achieve the largest variance
reductions, as both exploit detailed knowledge of the integrand's shape.  Stratified sampling and
antithetic sampling provide moderate reductions with minimal implementation overhead.  The
Beta$(2,1)$ importance sampling density is poorly matched to the integrand and fails to improve
upon simple Monte Carlo, illustrating that a badly chosen proposal can be just as costly as no
variance reduction technique at all.

\newpage

# Problem 4: Project Topic Report

## Introduction

This project investigates how multiplicative shear bias propagates through a 
weak gravitational lensing mass reconstruction pipeline using standard methods. 
Weak lensing is among the most powerful probes of cosmological structure. 
By measuring the coherent distortion of background galaxy shapes induced by 
intervening mass distributions, one can reconstruct the projected surface mass density 
$\kappa(\boldsymbol{\theta})$ of galaxy clusters and large-scale structure. 
The standard pipeline consists of two stages: a shape measurement stage, in which 
galaxy ellipticities are estimated from noisy, PSF-convolved images using a method 
such as NGmix with metacalibration, followed by a mass reconstruction stage, in which 
the estimated shear field is inverted to recover $\kappa$ using a method such as Kaiser-Squires. 
Both stages introduce systematic errors, and understanding how errors in the first stage propagate 
into errors in the second is a fundamental question in weak lensing statistics with direct implications 
for survey design. This project addresses that question through a controlled Monte Carlo simulation study 
using GalSim for image generation, NGmix with metacalibration for shape measurement, and Kaiser-Squires for 
mass reconstruction, applying variance reduction techniques and bootstrap inference drawn from the MA 551 
curriculum to a problem not previously addressed in this form.

## Problem Description

All practical shear estimators, including NGmix with metacalibration, produce
output carrying a multiplicative bias $m$ and additive bias $c$ of the form
$$
\hat{g}_i = (1 + m)\,g_i^{\mathrm{true}} + c, \qquad i = 1, 2,
$$
where $m$ and $c$ depend on signal-to-noise ratio (SNR) and galaxy size relative
to the PSF.  Metacalibration partially corrects for this bias by applying small
artificial shears to each image and numerically computing the shear response
matrix $\mathbf{R}$, but the correction is imperfect: residual bias remains,
and the response matrix is itself estimated with uncertainty from finite galaxy
samples.  Downstream, Kaiser-Squires mass reconstruction recovers $\kappa$ from
the estimated shear via a linear FFT-based inversion of the lensing kernel.
Because the inversion is linear, multiplicative bias in the shear propagates
directly into the recovered $\kappa$, but the precise relationship is modulated
by the smoothing kernel, boundary effects, and the shape noise floor.

Shape noise, the intrinsic ellipticity dispersion of source galaxies, is an
irreducible source of error in any weak lensing measurement: it does not
diminish with larger datasets at the per-pixel level, and the pixel-wise
variance of the reconstructed $\kappa$ map is proportional to
$\sigma_\epsilon^2$ divided by the effective galaxy number density.  This study
fixes shape noise at a realistic level ($\sigma_\epsilon = 0.3$) and treats it
as a constant baseline, investigating how residual multiplicative bias $m$
contributes to reconstruction error above that floor.  Sky background noise,
by contrast, varies strongly with observing conditions; following the regime of
balloon-borne and space-based observatories such as SuperBIT, which operate
above the bulk of the atmosphere where sky background is negligible relative to
detector read noise and galaxy Poisson noise, we treat sky noise as small and
fix the noise model accordingly.  These two simplifications reduce the parameter
space to $(m, \mathrm{SNR})$, where SNR is set by galaxy flux and detector
noise, while still capturing the dominant noise sources that interact with bias
in the reconstruction.  Characterising the reconstruction error surface
$\mathcal{E}(m, \mathrm{SNR})$ in this regime requires a controlled Monte Carlo
study, since the joint effects of shape noise, PSF convolution, metacalibration
response uncertainty, and KS inversion do not admit a simple closed-form
treatment.  This characterisation is a prerequisite for setting the bias
tolerance that shape measurement pipelines must meet, with direct consequences
for survey exposure time and catalog size requirements.

## Objectives

The goals of the project are as follows.  First, to characterise the
reconstruction error surface $\mathcal{E}(m, \mathrm{SNR})$ via Monte Carlo
simulation: for a grid of controlled multiplicative bias values and SNR levels,
measure the normalised $L^2$ error $\|\kappa_{\mathrm{rec}} -
\kappa_{\mathrm{true}}\|/\|\kappa_{\mathrm{true}}\|$ in the Kaiser-Squires
reconstruction across many independent noise realisations, with shape noise
held fixed at a realistic level throughout.  Second, to apply stratified
sampling over the $(m, \mathrm{SNR})$ parameter grid and importance sampling
over the galaxy ellipticity distribution to estimate the bias-integrated
reconstruction error efficiently, and to compare variance reduction against
naive uniform grid sampling.  Third, to assess how bootstrap confidence
intervals behave in the presence of systematically biased shear input: for each
grid cell, percentile and BCa intervals will be constructed for a scalar summary
statistic of the reconstruction, specifically the peak amplitude of
$\kappa_{\mathrm{rec}}$ and the total aperture mass within a fixed angular
radius, and empirical coverage will be assessed against the nominal $95\%$
level, with particular attention to whether coverage degrades as a function of
$m$ and whether BCa intervals are more robust to bias than percentile
intervals.  Fourth, to conduct a power analysis determining the
minimum number of simulated galaxy fields required to measure the multiplicative
bias $m$ with sufficient precision to determine whether it falls within the
acceptable tolerance, connecting the statistical result directly to a concrete
survey design specification.

## Approach and Methodology

Galaxy images will be simulated using GalSim with Gaussian galaxy profiles
convolved with a Gaussian PSF and Gaussian read noise, parameterised by a grid
of SNR values at a fixed size ratio $r = \sigma_{\mathrm{gal}}/\sigma_{\mathrm{PSF}} = 1$,
which is representative of high-resolution space imaging conditions.  Gaussian
profiles are used in place of more complex morphologies as a standard
proof-of-concept simplification appropriate for a bias propagation study; the
goal is to characterise how bias propagates statistically rather than to produce
a definitive calibration for a specific survey.  A Gaussian convergence profile
$\kappa_{\mathrm{true}}(\boldsymbol{\theta}) =
\exp(-|\boldsymbol{\theta}|^2 / 2\sigma_\ell^2)$ with $\sigma_\ell =
0.5\,\mathrm{arcmin}$ serves as the ground truth, with an NFW profile used as
a secondary test case.  The true shear field $(\gamma_1, \gamma_2)$ is
computed analytically from $\kappa_{\mathrm{true}}$ via the lensing kernel.
Galaxy ellipticities are drawn from a Gaussian intrinsic shape distribution with
dispersion $\sigma_\epsilon = 0.3$ per component before PSF convolution and
noise addition.  Sky background is set to a level consistent with
balloon/space-based observing conditions, where it is negligible relative to
detector and Poisson noise sources.

Shape measurement will be performed using NGmix with metacalibration.  To
enable controlled injection of known bias, the study will operate in two modes.
In the \emph{natural bias} mode, NGmix is applied directly to simulated images
and its residual multiplicative bias $m$ is measured empirically as a function
of SNR and size ratio by linear regression of estimated shear on true shear
across Monte Carlo replicates.  In the \emph{injected bias} mode, the true
shear field is artificially scaled by $(1+m)$ for a range of fixed $m$ values
before passing to Kaiser-Squires, allowing precise control over the input bias
independent of the shape measurement step.  This two-mode design makes the
simulation pipeline modular: the shape measurement and reconstruction stages
are decoupled so that alternative estimators can be substituted in future work
without restructuring the statistical analysis.

Kaiser-Squires mass reconstruction is applied to the estimated shear field on
a regular grid with a fixed Gaussian smoothing kernel.  The normalised $L^2$
reconstruction error $\mathcal{E}$ is computed on the interior of the field to
avoid boundary artefacts.

Statistical analysis will comprise four components.  For the reconstruction
error surface, mean and standard deviation of $\mathcal{E}$ will be estimated
across $B = 500$ noise realisations per grid cell, and a quadratic regression
surface $\mathcal{E}(m, \mathrm{SNR})$ will be fitted by weighted least squares
with jackknife standard errors on the regression coefficients.  For variance
reduction, stratified sampling will allocate replicates proportional to
within-stratum variance of $\mathcal{E}$, and an importance sampling
distribution concentrated in the high-$|m|$, low-SNR region will be compared
against uniform allocation across the grid.  For bootstrap coverage, percentile
and BCa confidence intervals will be compared across the bias grid to determine
whether standard resampling-based inference fails under systematic input bias
and, if so, at what level of $m$ the failure becomes statistically significant.
Finally, a power analysis will determine the minimum number of galaxy fields
$N_{\mathrm{fields}}$ required to constrain $m$ to within a target precision
at $5\sigma$ significance.  This analysis is conducted in the natural bias mode,
where $m$ is unknown and estimated by linear regression of NGmix shear output
on true shear across fields; the power curve gives the sample size at which
the regression-based estimate of $m$ is precise enough to determine whether it
falls within an acceptable tolerance.

\newpage

# Appendix: Session Information

```{r session_info}
sessionInfo()
```
