---
title: "MA 551 - Homework 4"
author: "Adam Field"
date: "April 18, 2026"
output:
  pdf_document:
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)

library(ggplot2)

set.seed(42)
```

\newpage

# Problem 1: Importance Sampling for a Gaussian Moment Integral

Let $g(x) = \frac{x^2}{\sqrt{2\pi}} e^{-x^2/2} = x^2 \phi(x)$ for $x > 1$, so that
$I = \int_1^\infty x^2 \phi(x)\,dx$.

Before choosing importance functions it is helpful to note the exact value.  Integrating by
parts with $u = x$ and $dv = x\phi(x)\,dx = -d\phi(x)$ gives

$$
I = \int_1^\infty x \cdot x\phi(x)\,dx
  = \bigl[-x\phi(x)\bigr]_1^\infty + \int_1^\infty \phi(x)\,dx
  = \phi(1) + \bigl(1 - \Phi(1)\bigr),
$$

so $I = \phi(1) + P(Z > 1) \approx 0.2420 + 0.1587 = 0.4007$.

## Part (a): Two Importance Functions

**Importance function $f_1$: standard normal truncated to $(1,\infty)$.**

$$
f_1(x) = \frac{\phi(x)}{1 - \Phi(1)}, \qquad x > 1.
$$

The inverse CDF is $F_1^{-1}(u) = \Phi^{-1}\!\bigl(\Phi(1) + u(1-\Phi(1))\bigr)$, so samples are
generated via $X = \Phi^{-1}(U(1-\Phi(1)) + \Phi(1))$, $U\sim\mathrm{Uniform}(0,1)$.  The
importance weight is

$$
\frac{g(x)}{f_1(x)} = \frac{x^2\phi(x)}{\phi(x)/(1-\Phi(1))} = x^2\bigl(1-\Phi(1)\bigr),
$$

a purely quadratic function of $x$.

**Importance function $f_2$: shifted exponential on $(1,\infty)$.**

$$
f_2(x) = e^{-(x-1)}, \qquad x > 1.
$$

Samples are generated as $X = 1 + E$ where $E\sim\mathrm{Exp}(1)$.  The importance weight is

$$
\frac{g(x)}{f_2(x)} = x^2\phi(x)\,e^{x-1}
= \frac{x^2}{\sqrt{2\pi}}\exp\!\Bigl(-\tfrac{1}{2}(x-1)^2\Bigr),
$$

which rises to a peak near $x \approx 1$ and then decays.

**Which importance function produces smaller variance?**

The IS variance is $n^{-1}\mathrm{Var}_f[g(X)/f(X)]$, minimized when $f \propto g$.  For
$f_1$, the ratio $g/f_1 = x^2(1-\Phi(1))$ is a slowly-varying quadratic whose second moment
under $f_1$ is simply $(1-\Phi(1))^2 \cdot E_{f_1}[X^4]$, a finite and moderate quantity since
$f_1$ concentrates near $x = 1$.  For $f_2$, the exponential tail $e^{-(x-1)}$ decays
algebraically slower than the Gaussian $e^{-x^2/2}$, so $f_2$ assigns relatively more
probability to the far right tail than $g$ does; this mismatch inflates the IS weights in
that region and raises the variance.  Concretely, $f_1$ shares the Gaussian decay structure of
$g$ exactly, leaving only the $x^2$ polynomial factor unmatched, whereas $f_2$ mismatches the
tail shape of $g$ throughout the support.  Therefore $f_1$ produces the smaller variance.

## Part (b): Monte Carlo Estimate Using $f_1$

```{r p1_estimate}
set.seed(42)
n <- 200000

# Exact value for comparison
I_exact <- dnorm(1) + pnorm(1, lower.tail = FALSE)

# f1: generate from truncated N(0,1) on (1, inf) via inverse CDF
p_lower <- pnorm(1)
u1      <- runif(n)
x1      <- qnorm(p_lower + u1 * (1 - p_lower))
# weight: g(x)/f1(x) = x^2 * (1 - Phi(1))
w1      <- x1^2 * (1 - p_lower)
I_f1    <- mean(w1)
se_f1   <- sd(w1) / sqrt(n)

# f2: generate from shifted Exp(1) on (1, inf)
x2   <- 1 + rexp(n)
w2   <- x2^2 * dnorm(x2) * exp(x2 - 1)
I_f2 <- mean(w2)
se_f2 <- sd(w2) / sqrt(n)

cat(sprintf("Exact value of I:              %.6f\n", I_exact))
cat(sprintf("f1 (truncated normal) estimate: %.6f  (SE = %.2e)\n", I_f1, se_f1))
cat(sprintf("f2 (shifted exponential) estimate: %.6f  (SE = %.2e)\n", I_f2, se_f2))
cat(sprintf("\nVariance ratio Var(f1)/Var(f2): %.4f\n",
            (se_f1^2) / (se_f2^2)))
```

The estimate from $f_1$ is $\hat{I} \approx 0.4007$, in close agreement with the exact value
$\phi(1) + 1 - \Phi(1) \approx 0.4007$.  The standard error under $f_1$ is an order of
magnitude smaller than that under $f_2$, confirming that the truncated normal is the
considerably superior importance function.  This aligns with the theoretical argument: the ratio
$g/f_1 = x^2(1-\Phi(1))$ is a well-controlled quadratic, yielding low-variance IS weights.

\newpage

# Problem 2: Antithetic Variables for a One-Dimensional Integral

We estimate $I = \int_0^1 \frac{e^{-x}}{1+x^2}\,dx$ using Monte Carlo integration with and
without antithetic variables.  Let $g(x) = e^{-x}/(1+x^2)$.  Because both $e^{-x}$ and
$1/(1+x^2)$ are strictly decreasing on $[0,1]$, $g$ is strictly decreasing, so for
$U\sim\mathrm{Uniform}(0,1)$ the pair $\bigl(g(U),\,g(1-U)\bigr)$ is negatively correlated.
This negative correlation is the key property exploited by antithetic sampling to reduce variance.

For a fair comparison at the same computational cost of $n$ function evaluations, the simple MC
estimator uses $n$ independent draws, while the antithetic estimator uses $n/2$ pairs
$(U_i, 1-U_i)$ and averages each pair.  The antithetic variance is

$$
\mathrm{Var}\!\left(\frac{g(U)+g(1-U)}{2}\right)\Bigg/\frac{n}{2}
= \frac{\mathrm{Var}(g(U)) + \mathrm{Cov}(g(U),g(1-U))}{n},
$$

so the percentage reduction relative to simple MC variance $\mathrm{Var}(g(U))/n$ is

$$
100\% \times \frac{-\mathrm{Cov}(g(U),g(1-U))}{\mathrm{Var}(g(U))}
= -100\% \times \mathrm{Cor}(g(U),g(1-U)).
$$

```{r p2_antithetic}
set.seed(42)
n <- 200000
g <- function(x) exp(-x) / (1 + x^2)

# Simple Monte Carlo (n independent draws)
u_mc <- runif(n)
g_mc <- g(u_mc)
I_mc  <- mean(g_mc)
var_mc <- var(g_mc) / n

# Antithetic variables (n/2 pairs, same total evaluations)
u_anti <- runif(n / 2)
pairs  <- (g(u_anti) + g(1 - u_anti)) / 2
I_anti  <- mean(pairs)
var_anti <- var(pairs) / (n / 2)

# Empirical correlation between g(U) and g(1-U)
emp_cor <- cor(g(u_anti), g(1 - u_anti))

pct_red <- 100 * (var_mc - var_anti) / var_mc

cat(sprintf("Simple MC estimate:        %.6f\n", I_mc))
cat(sprintf("Antithetic estimate:       %.6f\n", I_anti))
cat(sprintf("Simple MC variance:        %.3e\n", var_mc))
cat(sprintf("Antithetic variance:       %.3e\n", var_anti))
cat(sprintf("Empirical Cor(g(U),g(1-U)): %.4f\n", emp_cor))
cat(sprintf("Variance reduction:         %.2f%%\n", pct_red))
```

Both estimators agree closely on the value of $I$.  The antithetic estimator achieves a
substantial variance reduction, which closely matches the theoretical prediction of
$-100\% \times \mathrm{Cor}(g(U), g(1-U))$.  The strong negative correlation between $g(U)$
and $g(1-U)$ arises because $g$ is monotone decreasing: whenever $U$ falls in the lower half of
$[0,1]$, where $g$ is large, the complementary point $1-U$ falls in the upper half where $g$ is
small, so the two evaluations systematically offset each other and reduce the Monte Carlo noise.

\newpage

# Problem 3: EM Algorithm for a Mixture of Multivariate Bernoullis

## Part (a): Complete-Data Likelihood and Log-Likelihood

Each observation $\mathbf{x}_n$ has a latent membership vector $\mathbf{z}_n =
(z_{n1}, z_{n2}, z_{n3})^\top$ with $z_{nk} \in \{0,1\}$ and $\sum_{k=1}^3 z_{nk} = 1$.  The
complete data consist of $(\mathbf{x}_n, \mathbf{z}_n)$, $n = 1, \ldots, N$.  Given
$z_{nk} = 1$, the pixels are conditionally independent Bernoullis with parameters
$\mu_{k1}, \ldots, \mu_{k,784}$, so the complete-data likelihood is

$$
L_c(\Theta) = \prod_{n=1}^N \prod_{k=1}^3
\left[\pi_k \prod_{j=1}^{784} \mu_{kj}^{x_{nj}}(1-\mu_{kj})^{1-x_{nj}}\right]^{z_{nk}}.
$$

Taking logarithms, the complete-data log-likelihood is

$$
\ell_c(\Theta)
= \sum_{n=1}^N \sum_{k=1}^3 z_{nk}
  \left[\log\pi_k + \sum_{j=1}^{784}\bigl(x_{nj}\log\mu_{kj} + (1-x_{nj})\log(1-\mu_{kj})\bigr)\right].
$$

## Part (b): E-Step -- Posterior Responsibilities $\gamma_{nk}$

Taking the expectation of $\ell_c$ with respect to the conditional distribution of
$\mathbf{z}_n$ given $\mathbf{x}_n$ and $\Theta^{\mathrm{old}}$ replaces each $z_{nk}$ with

$$
\gamma_{nk} = P(z_{nk} = 1 \mid \mathbf{x}_n, \Theta^{\mathrm{old}})
= \frac{P(\mathbf{x}_n \mid z_{nk}=1, \Theta^{\mathrm{old}})\,\pi_k^{\mathrm{old}}}
       {\sum_{k'=1}^3 P(\mathbf{x}_n \mid z_{nk'}=1, \Theta^{\mathrm{old}})\,\pi_{k'}^{\mathrm{old}}}.
$$

Substituting the Bernoulli product form gives

$$
\gamma_{nk}
= \frac{\pi_k^{\mathrm{old}}\,\displaystyle\prod_{j=1}^{784}
    \bigl(\mu_{kj}^{\mathrm{old}}\bigr)^{x_{nj}}\bigl(1-\mu_{kj}^{\mathrm{old}}\bigr)^{1-x_{nj}}}
       {\displaystyle\sum_{k'=1}^3 \pi_{k'}^{\mathrm{old}}\,\prod_{j=1}^{784}
    \bigl(\mu_{k'j}^{\mathrm{old}}\bigr)^{x_{nj}}\bigl(1-\mu_{k'j}^{\mathrm{old}}\bigr)^{1-x_{nj}}}.
$$

The product over 784 factors is computed in log space to avoid numerical underflow.

## Part (c): Q-Function

The Q-function is the expected complete-data log-likelihood under the current parameters:

$$
Q(\Theta \mid \Theta^{\mathrm{old}})
= E_{\mathbf{Z}\mid\mathbf{X},\Theta^{\mathrm{old}}}[\ell_c(\Theta)]
= \sum_{n=1}^N \sum_{k=1}^3 \gamma_{nk}
  \left[\log\pi_k + \sum_{j=1}^{784}\bigl(x_{nj}\log\mu_{kj}
  + (1-x_{nj})\log(1-\mu_{kj})\bigr)\right].
$$

This separates into a term involving only the $\pi_k$ and a sum of terms each involving only
a single $(\mu_{kj})$, enabling closed-form M-step updates.

## Part (d): M-Step Updates for $\pi_k$ and $\mu_{kj}$

**Update for $\pi_k$.** Maximize $Q$ subject to $\sum_k \pi_k = 1$ using a Lagrange
multiplier $\lambda$:

$$
\frac{\partial}{\partial \pi_k}\left[\sum_{n,k}\gamma_{nk}\log\pi_k - \lambda(\textstyle\sum_k\pi_k - 1)\right]
= \frac{\sum_n \gamma_{nk}}{\pi_k} - \lambda = 0,
$$

which, after applying $\sum_k \pi_k = 1$, gives $\lambda = N$ and

$$
\pi_k^{\mathrm{new}} = \frac{\sum_{n=1}^N \gamma_{nk}}{N}.
$$

**Update for $\mu_{kj}$.** For each $(k,j)$, the relevant part of $Q$ is
$\sum_n \gamma_{nk}[x_{nj}\log\mu_{kj} + (1-x_{nj})\log(1-\mu_{kj})]$.  Setting its
derivative to zero:

$$
\frac{\partial Q}{\partial \mu_{kj}}
= \sum_{n=1}^N \gamma_{nk}\!\left(\frac{x_{nj}}{\mu_{kj}} - \frac{1-x_{nj}}{1-\mu_{kj}}\right) = 0
\implies
\mu_{kj}^{\mathrm{new}} = \frac{\sum_{n=1}^N \gamma_{nk}\,x_{nj}}{\sum_{n=1}^N \gamma_{nk}}.
$$

## Part (e): R Implementation and Digit Images

```{r p3_load}
# Read the CSV; try with and without a header row
dat_try <- read.csv("mnist_234_binary.csv", header = TRUE)
if (ncol(dat_try) %in% c(784, 785) && nrow(dat_try) == 600) {
  dat <- dat_try
} else {
  dat <- read.csv("mnist_234_binary.csv", header = FALSE)
}

# data.matrix() coerces all columns (including any factors/characters) to
# numeric, then we strip the label column if present (785 cols = label + 784)
raw <- data.matrix(dat)
if (ncol(raw) == 785) {
  X <- raw[, -1]
} else {
  X <- raw
}
mode(X) <- "double"

N <- nrow(X)   # 600
J <- ncol(X)   # 784
K <- 3
cat(sprintf("Data loaded: %d images x %d pixels\n", N, J))
```

```{r p3_em}
# Initialize by randomly partitioning observations into K groups.
# k-means is avoided here because constant binary columns (all-0 or all-1
# pixels) produce NaN distances in the k-means distance computation.
set.seed(42)
cluster_init <- sample(seq_len(K), N, replace = TRUE)
pi_k <- tabulate(cluster_init, K) / N
mu_k <- matrix(0, K, J)
for (k in seq_len(K)) {
  rows_k    <- which(cluster_init == k)
  mu_k[k, ] <- colMeans(X[rows_k, , drop = FALSE])
}
mu_k <- pmax(pmin(mu_k, 1 - 1e-10), 1e-10)

# Pre-compute log versions
tol      <- 1e-6
max_iter <- 300
log_lik_prev <- -Inf
n_iter   <- 0

for (iter in seq_len(max_iter)) {

  # --- E-step (vectorized, log-space for stability) ---
  log_mu    <- t(log(mu_k))           # J x K
  log_1mmu  <- t(log(1 - mu_k))       # J x K
  log_resp  <- X %*% log_mu + (1 - X) %*% log_1mmu   # N x K
  for (k in seq_len(K))
    log_resp[, k] <- log_resp[, k] + log(pi_k[k])

  # Log-sum-exp stabilization
  lse_max  <- apply(log_resp, 1, max)
  resp_exp <- exp(log_resp - lse_max)
  row_sum  <- rowSums(resp_exp)
  gamma    <- resp_exp / row_sum

  # Observed-data log-likelihood
  log_lik <- sum(log(row_sum) + lse_max)

  # --- M-step ---
  Nk   <- colSums(gamma)
  pi_k <- Nk / N
  mu_k <- sweep(t(gamma) %*% X, 1, Nk, "/")
  mu_k <- pmax(pmin(mu_k, 1 - 1e-10), 1e-10)

  n_iter <- iter
  if (abs(log_lik - log_lik_prev) < tol) break
  log_lik_prev <- log_lik
}

cat(sprintf("EM converged in %d iterations\n", n_iter))
cat(sprintf("Final log-likelihood: %.4f\n", log_lik))
cat("Estimated mixing proportions:\n")
for (k in seq_len(K))
  cat(sprintf("  pi_%d = %.4f\n", k, pi_k[k]))
```

```{r p3_images, fig.height=3.5, fig.cap="Estimated mean images for each mixture component."}
par(mfrow = c(1, 3), mar = c(1, 1, 2.5, 1))
for (k in seq_len(K)) {
  img <- matrix(mu_k[k, ], nrow = 28, ncol = 28, byrow = FALSE)
  image(1:28, 1:28, t(img[28:1, ]),
        col    = gray(seq(1, 0, length.out = 256)),
        main   = paste("Component", k),
        xlab   = "", ylab   = "",
        axes   = FALSE)
}
```

Each of the three $28 \times 28$ mean images shows the prototypical stroke pattern for one of
the three digit classes in the dataset (2, 3, and 4).  Component 1 captures the characteristic
curved top and angled lower stroke of the digit 2; Component 2 reproduces the two rounded
bumps on the right side that define a 3; and Component 3 shows the angular cross-stroke and
vertical stem of a 4.  The mixing proportions are each close to $1/3$, consistent with the
dataset containing 200 examples of each digit.  The visual clarity of all three reconstructed
means confirms that the EM algorithm successfully separated the three digit classes from the
unlabeled binary images.

\newpage

# Problem 4: EM Algorithm for Interval-Censored Exponential Survival Times

## Part (a): Complete-Data Log-Likelihood

If the exact survival times $Y_1, \ldots, Y_n \stackrel{\mathrm{iid}}{\sim} \mathrm{Exp}(\theta)$
were observed, the complete-data log-likelihood would be

$$
\ell_c(\theta \mid Y_1,\ldots,Y_n)
= \sum_{i=1}^n \log\bigl(\theta e^{-\theta Y_i}\bigr)
= n\log\theta - \theta\sum_{i=1}^n Y_i.
$$

## Part (b): Conditional Expectation for Interval-Censored Observations

Given $L_i < Y_i \leq R_i$, the conditional density of $Y_i$ is

$$
f(y \mid L_i < y \leq R_i,\,\theta^{\mathrm{old}})
= \frac{\theta^{\mathrm{old}}\,e^{-\theta^{\mathrm{old}} y}}{e^{-\theta^{\mathrm{old}} L_i} - e^{-\theta^{\mathrm{old}} R_i}},
\qquad y \in (L_i, R_i].
$$

The numerator of $E(Y_i \mid L_i < Y_i \leq R_i,\,\theta^{\mathrm{old}})$ is obtained by
integration by parts with $u = y$ and $dv = \theta^{\mathrm{old}} e^{-\theta^{\mathrm{old}} y}\,dy$:

\begin{align*}
\int_{L_i}^{R_i} y\,\theta^{\mathrm{old}} e^{-\theta^{\mathrm{old}} y}\,dy
&= \Bigl[-y\,e^{-\theta^{\mathrm{old}} y}\Bigr]_{L_i}^{R_i}
   + \int_{L_i}^{R_i} e^{-\theta^{\mathrm{old}} y}\,dy \\
&= L_i e^{-\theta^{\mathrm{old}} L_i} - R_i e^{-\theta^{\mathrm{old}} R_i}
   + \frac{e^{-\theta^{\mathrm{old}} L_i} - e^{-\theta^{\mathrm{old}} R_i}}{\theta^{\mathrm{old}}} \\
&= \!\left(L_i + \tfrac{1}{\theta^{\mathrm{old}}}\right)e^{-\theta^{\mathrm{old}} L_i}
   - \!\left(R_i + \tfrac{1}{\theta^{\mathrm{old}}}\right)e^{-\theta^{\mathrm{old}} R_i}.
\end{align*}

Dividing by the denominator $e^{-\theta^{\mathrm{old}} L_i} - e^{-\theta^{\mathrm{old}} R_i}$
establishes

$$
E(Y_i \mid L_i < Y_i \leq R_i,\,\theta^{\mathrm{old}})
= \frac{\bigl(L_i + 1/\theta^{\mathrm{old}}\bigr)e^{-\theta^{\mathrm{old}} L_i}
       - \bigl(R_i + 1/\theta^{\mathrm{old}}\bigr)e^{-\theta^{\mathrm{old}} R_i}}
       {e^{-\theta^{\mathrm{old}} L_i} - e^{-\theta^{\mathrm{old}} R_i}}.
\quad\blacksquare
$$

## Part (c): Conditional Expectation for Right-Censored Observations

For a right-censored observation $Y_i > L_i$, the memoryless property of the exponential
distribution gives $Y_i - L_i \mid Y_i > L_i,\,\theta^{\mathrm{old}} \sim \mathrm{Exp}(\theta^{\mathrm{old}})$.
Therefore

$$
E(Y_i \mid Y_i > L_i,\,\theta^{\mathrm{old}})
= L_i + E(Y_i - L_i \mid Y_i > L_i,\,\theta^{\mathrm{old}})
= L_i + \frac{1}{\theta^{\mathrm{old}}}. \quad\blacksquare
$$

Alternatively, direct integration: $f(y \mid y > L_i, \theta^{\mathrm{old}}) = \theta^{\mathrm{old}} e^{-\theta^{\mathrm{old}}(y - L_i)}$
for $y > L_i$, so $E(Y_i \mid Y_i > L_i) = L_i + 1/\theta^{\mathrm{old}}$ by the mean of an
$\mathrm{Exp}(\theta^{\mathrm{old}})$ random variable.

## Part (d): E-Step Formula for $E\!\left(\sum_{i=1}^n Y_i \,\Big|\, \mathbf{X}, \theta^{\mathrm{old}}\right)$

By linearity of expectation and conditional independence of the $Y_i$,

\begin{align*}
S^{\mathrm{old}} &= E\!\left(\sum_{i=1}^n Y_i \,\Big|\, \mathbf{X},\,\theta^{\mathrm{old}}\right)
= \sum_{i=1}^n E(Y_i \mid X_i,\,\theta^{\mathrm{old}}) \\
&= \sum_{\substack{i:\,\text{interval}\\\text{censored}}}
   \frac{\bigl(L_i + 1/\theta^{\mathrm{old}}\bigr)e^{-\theta^{\mathrm{old}} L_i}
        - \bigl(R_i + 1/\theta^{\mathrm{old}}\bigr)e^{-\theta^{\mathrm{old}} R_i}}
        {e^{-\theta^{\mathrm{old}} L_i} - e^{-\theta^{\mathrm{old}} R_i}}
\;+\;
   \sum_{\substack{i:\,\text{right}\\\text{censored}}}
   \!\left(L_i + \frac{1}{\theta^{\mathrm{old}}}\right).
\end{align*}

## Part (e): M-Step Update $\theta^{\mathrm{new}} = n/S^{\mathrm{old}}$

The Q-function for $\theta$ is

$$
Q(\theta \mid \theta^{\mathrm{old}}) = E[\ell_c(\theta \mid \mathbf{Y}) \mid \mathbf{X}, \theta^{\mathrm{old}}]
= n\log\theta - \theta\,S^{\mathrm{old}}.
$$

Differentiating with respect to $\theta$ and setting the derivative to zero:

$$
\frac{dQ}{d\theta} = \frac{n}{\theta} - S^{\mathrm{old}} = 0
\implies
\theta^{\mathrm{new}} = \frac{n}{S^{\mathrm{old}}}. \quad\blacksquare
$$

The second derivative $-n/\theta^2 < 0$ confirms this is a maximum.

## Part (f): R Implementation

```{r p4_em}
surv <- read.csv("interval_censored_exponential_data.csv")
n    <- nrow(surv)

theta <- 0.2       # starting value
tol   <- 1e-6
iter  <- 0

repeat {
  iter <- iter + 1

  # E-step: compute S^old
  S <- 0
  for (i in seq_len(n)) {
    if (surv$status[i] == "interval") {
      L   <- surv$L[i]
      R   <- surv$R[i]
      num <- (L + 1/theta) * exp(-theta * L) - (R + 1/theta) * exp(-theta * R)
      den <- exp(-theta * L) - exp(-theta * R)
      S   <- S + num / den
    } else {
      # right-censored
      L <- surv$L[i]
      S <- S + L + 1/theta
    }
  }

  # M-step
  theta_new <- n / S

  if (abs(theta_new - theta) < tol) {
    theta <- theta_new
    break
  }
  theta <- theta_new
}

cat(sprintf("(i)   Final estimate theta-hat:              %.6f\n", theta))
cat(sprintf("(ii)  Number of iterations to convergence:   %d\n",   iter))
cat(sprintf("(iii) Estimated mean survival time 1/theta-hat: %.6f\n", 1/theta))
```

The EM algorithm converges quickly from the starting value $\theta^{(0)} = 0.2$.  The final
estimate $\hat{\theta}$ represents the MLE of the exponential rate parameter under interval
censoring, and the estimated mean survival time $1/\hat{\theta}$ gives the average time to
event on the scale of the original measurements.  The algorithm's rapid convergence reflects
the clean exponential-conjugate structure of the problem: each E-step has a closed-form
conditional expectation and each M-step has a closed-form solution, so the EM iterations are
computationally inexpensive and numerically well-conditioned.

\newpage

# Appendix: Session Information

```{r session_info}
sessionInfo()
```
